---
title: "Growth curves"
output: html_notebook
editor_options:
  code_folding: hide
  chunk_output_type: inline
---
Growth curves for $\Delta$6$-\Delta$SpoIIE and $\Delta$6-WT
```{r, message=FALSE}
library(here)
library(tidyverse)
library(lubridate)
library(broom)
library(cowplot)
```

## Read data from plate reade file  

```{r, message=FALSE}
# read this file
xpt.file <- here("data", "growth-curve", "OD600_96well_16hr_37dC_shake_20200305.txt")

# number of times plate was read
xpt.reads = grep("Reads", readLines(xpt.file), value = T)%>%
  str_remove(pattern = ".*,") %>% parse_number()

# start line of OD table
start.line = grep("^Time.*A1", readLines(xpt.file))

# end line of OD table
aprox.end.line.n= floor(start.line+xpt.reads)-start.line

# read data and organize it
input <- 
  read_tsv(xpt.file,
       skip = start.line-1,
       n_max = aprox.end.line.n-1,
       col_types = cols(Time = "c"))%>%
  # rename temp column
    rename("temp"=2)%>%
  # Convert time to hms
    mutate(Time=lubridate::hms(Time))%>%
    mutate(Time=as.duration(Time))%>%
  # remove empty lines
    filter(Time != 0)

```

In these files there are reads evry 2 min. I will avearge on window of 8 reads.
```{r, message=FALSE}
# window size
win <- 8

#number of points after averaging
n.group <-floor( nrow(input)/win)

# remove last rows so it divides by win
input <- input[1:(win*n.group),]

input.avg <- 
  input%>%
  #add grouping variable 
    mutate(g=ntile(1:(win*n.group),n.group))%>%
    group_by(g)%>%
    summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))%>%
  #remove grouping var
    select(-g) %>%
  # convert time to hours
    mutate(Time = Time/3600)

#trnaform to long format
d <- gather(input.avg, key="well", value = "OD600", colnames(input)[-c(1:2)])

#add metadata
meta <- read_csv(here("data", "growth-curve","20200305_spoIIE_d6_layout.csv"))
d <-merge(d, meta)
```

### plotting the data

Plot by well
```{r, message=FALSE}
d%>%
  ggplot(aes(x=Time, y=OD600))+
  geom_line(aes(color=interaction(strain,colony, media)))+
  theme_cowplot()+
  facet_grid(row~col)+
  theme(legend.position = "none")


```

A few wells were ouliers and need to be filtered out:  
* coloumn 6 and 11 well got lower volume of input due to protocol error.
* Wells B2 and D8 probably had some particulate matter  
* blanks of A7 and H8, seem contaminated  

```{r, message=FALSE}

outlier.wells <-
  meta%>%
  filter(col==6|col==11|well=="D8"|well=="A7"|well=="H8"|well=="B2")%>%
  select(well)%>%
  deframe()

d <- d%>%
  filter(!(well%in%outlier.wells))
  
```

Verify that initial OD was simmmilar across strains and replicates
```{r, message=FALSE}

d%>%  
  filter(Time==min(d$Time))%>%
  ggplot(aes(x=colony, y=OD600))+
  geom_boxplot()+
  geom_jitter(width = .2, shape=21, size=2, fill= "white")+
  theme_cowplot()+
    panel_border()+
    facet_grid(media ~ strain, scales = "free")+
  ylim(0,NA)

```


### Plot by strain and media  

```{r, message=FALSE}
d%>%
  filter(strain!="blank")%>%
  ggplot(aes(x=Time, y=OD600))+
      geom_line(aes(group=well,color=colony))+
      theme_cowplot()+
      facet_grid(media~strain)


```

The replicate colonies are on top of each other.How about the different strains?  
```{r}
d%>%
  filter(strain!="blank")%>%
  ggplot(aes(x=Time, y=OD600))+
      geom_line(aes(group=well,color=strain, linetype=colony))+
      theme_cowplot()+
      facet_grid(media~.)

```

For the growth phase they look exactly the same.  

## Focus on time of growth

```{r, message=FALSE}
d%>% 
  filter(strain!="blank")%>%
  filter(Time<9.5)%>%
  group_by(well)%>%
    ggplot(aes(x=Time, y=OD600))+
      geom_line(aes(color=colony, linetype=rep))+
      theme_cowplot()+
      facet_grid(media~strain)+
  coord_cartesian(ylim=c(0,1))

```

## Growth Rates 
I am using a seperate program to estimate growth rate. The program "GrowthRates" is described in Hall et al. (2013). *Growth rates made easy*. Molecular biology and evolution, 31(1), 232-238. I downloaded the executables from https://sourceforge.net/projects/growthrates/ .  
I will export the averaged data in a format suitable for GrowthRates: First column in minutes, and then the data, including a single blank well. I will also seperate LB from DSM, each with its own blank.  
```{r, eval=FALSE,echo=FALSE}
Min <- input.avg%>%
  filter(Time<9.5)%>%
  select(Time)%>%
  deframe()
Min <- round(Min*24, digits = 2)

#LB
export.meta <- 
  meta%>%
    filter(!(well%in%outlier.wells))%>%
    filter(media=="LB")

export <-     
input.avg%>%
  filter(Time<9.5)%>%
  select(export.meta$well[export.meta$strain!="blank"])

blank <- input.avg%>%
  filter(Time<9.5)%>%
  select(export.meta$well[export.meta$strain=="blank"])
blank <- blank[1]

write_tsv(cbind (Min,blank,export), file =  here("data/growth-curve","GrowthRates/LB.txt"))

#DSM
export.meta <- 
  meta%>%
    filter(!(well%in%outlier.wells))%>%
    filter(media=="DSM")

export <-     
input.avg%>%
  filter(Time<9.5)%>%
  select(export.meta$well[export.meta$strain!="blank"])

blank <- input.avg%>%
  filter(Time<9.5)%>%
  select(export.meta$well[export.meta$strain=="blank"])
blank <- blank[1]

write_tsv(cbind (Min,blank,export), file = here("data/growth-curve","GrowthRates/DSM.txt"))

```

I have run GrowthRates externally and will now import the results into R.  (GrowthRates commands: GrowthRates -i LB.txt -w 1; GrowthRates -i DSM.txt -w 1)

```{r, echo=FALSE, message=FALSE}
lb.gr <- read_tsv(here("data/growth-curve","GrowthRates/LB.summary"), skip=1)
dsm.gr <- read_tsv(here("data/growth-curve","GrowthRates/DSM.summary"), skip=1)

gr <- left_join(meta,rbind(lb.gr, dsm.gr), by=c("well"="Well"))

gr %>%
  filter(strain!="blank")%>%
  filter(!(well%in%outlier.wells))%>%
  # mutate(strain=fct_rev(strain))%>%
  ggplot(aes(x=strain, y=Rate))+
  geom_boxplot()+
  geom_jitter( aes(color=colony, shape=colony),size=3, height=0, width=0.3, fill="white")+
  theme_cowplot()+panel_border()+
  # coord_flip()+
  scale_shape_manual(values = c(21,22,24))+
  facet_wrap(~media, )+
    theme(axis.text.x = element_text(angle=90, vjust = 0.5))+
  ylim(0,NA)+
  ylab("Growth Rate")

```


```{r, echo=FALSE}
gr %>%
  filter(strain!="blank")%>%
  filter(!(well%in%outlier.wells))%>%
  ggplot(aes(x=strain, y=`lag time (minutes)`))+
  geom_boxplot()+
  geom_jitter( aes(color=colony, shape=colony),size=3, height=0, width=0.3, fill="white")+
  theme_cowplot()+panel_border()+
  # coord_flip()+
  scale_shape_manual(values = c(21,22,24))+
  facet_wrap(~media, )+
    theme(axis.text.x = element_text(angle=90, vjust = 0.5))+
  ylim(0,NA)
```

```{r, echo=FALSE}
gr %>%
  filter(strain!="blank")%>%
  filter(!(well%in%outlier.wells))%>%
  mutate(`Max OD`=as.numeric(`Max OD`))%>%
  ggplot(aes(x=strain, y=`Max OD`))+
  geom_boxplot()+
  geom_jitter( aes(color=colony, shape=colony),size=3, 
               height=0, width=0.3, fill="white")+
  theme_cowplot()+
  panel_border()+
  scale_shape_manual(values = c(21,22,24))+
  facet_wrap(~media )+
    theme(axis.text.x = element_text(angle=90, vjust = 0.5))+
  ylim(0,NA)

```


### ANOVA for LB  
```{r}
print("== Growth Rate ==")
gr%>%
  filter(strain!="blank")%>%
    filter(media=="LB")%>%
    filter(!(well%in%outlier.wells))%>%
  aov(Rate~strain+colony+rep, data=.)%>%
  summary()
print("== lag time (minutes) ==")
gr%>%
  filter(strain!="blank")%>%
    filter(media=="LB")%>%
  aov(`lag time (minutes)`~strain+colony+rep, data=.)%>%
  summary()
print("== Max OD ==")
gr%>%
  filter(strain!="blank")%>%
    filter(media=="LB")%>%
  mutate(`Max OD`=as.numeric(`Max OD`))%>%
  aov(`Max OD`~strain+colony+rep, data=.)%>%
  summary()
```

lag time (minutes) by strain post hoc  
```{r}
gr%>%
  filter(strain!="blank")%>%
    filter(media=="LB")-> tmp
t.test(tmp$`lag time (minutes)`~ tmp$strain)
```

 ### ANOVA for DSM  
```{r}
print("== Growth Rate ==")
gr%>%
  filter(strain!="blank")%>%
    filter(media=="DSM")%>%
    filter(!(well%in%outlier.wells))%>%
  aov(Rate~strain+colony+rep, data=.)%>%
  summary()
print("== lag time (minutes) ==")
gr%>%
  filter(strain!="blank")%>%
    filter(media=="DSM")%>%
  aov(`lag time (minutes)`~strain+colony+rep, data=.)%>%
  summary()
print("== Max OD ==")
gr%>%
  filter(strain!="blank")%>%
    filter(media=="DSM")%>%
  mutate(`Max OD`=as.numeric(`Max OD`))%>%
  aov(`Max OD`~strain+colony+rep, data=.)%>%
  summary()
```

Max OD by strain post hoc  

```{r}
gr%>%
  filter(strain!="blank")%>%
    filter(media=="DSM")-> tmp
t.test(as.numeric(tmp$`Max OD`)~ tmp$strain)
```

# **Take aways** 

## DSM
The plan for the co-evolution experiment is to use DSM. Significant strain dependent differences are seen for the **Max OD**. WT has a higher average Max OD, but the difference seems quite small. I should say that seeing a difference in this parameter is not surprising, as that is the point where the two strains are expected to behave differently: WT sporulates, while $\Delta$spoIIE cannot.

## LB  
Significant strain dependent differences are seen for the lag time, with WT having a longer lag period by nearly 5 minutes, on average. That seems like quite a lot. 



